### 2025-01-10 sections/2-methods.tex
added notation for tree and root

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e.,\,it is the point that minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e.,\,$C(c, r) \subset B(c, r)$.

@@

We start by performing a divisive hierarchical clustering on the dataset using CLAM.
The procedure is almost identical to that outlined in CHESS~\cite{ishaq2019clustered}, but with better selection of poles for partitioning (see Algorithm~\ref{alg:methods:partition}).
After building the tree, we also perform a depth-first reordering of the dataset (see Section~\ref{sec:methods:clustering:depth-first-reordering}).

For a cluster $C$ with $|C|$ points, we begin by taking a random subsample $S$ of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in $S$.
Using these distances, we compute the \textit{geometric median} of $S$; in other words, we find the point that minimizes the sum of distances to all other points in $S$.
We define the \textit{center} of $C$ to be this geometric median.

The \textit{radius} of $C$ is the maximum distance from the center to any other point in $C$.
The point that is responsible for that radius (i.e.,\,the furthest point from the center) is designated the \textit{left pole} and the point that is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster that are closer to the left pole than to the right pole, and the right child contains all points in the cluster that are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points that are equidistant from the two poles.
Starting from a root cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria, for example, minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc.
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

Note that Algorithm~\ref{alg:methods:partition} does not necessarily produce a balanced tree.
In fact, for real datasets, we expect anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g.,\,in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

@@

In CHESS, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Although this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree.
Assuming a balanced tree, and thus $\mathcal{O}(\log n)$ depth, this approach had a memory overhead of $\mathcal{O}(n \log n)$.
In this work, we introduce a new approach wherein, after building the cluster tree, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \textit{cardinality} and an \textit{offset} to access its points from the dataset, based on a useful invariant.
The root cluster has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.
With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers the same time complexity as CHESS with $\mathcal{O}(n)$ space.

@@

The asymptotic complexity of Partition is the same as described in~\cite{ishaq2019clustered}.
By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of partitioning and $\mathcal{O}(n \log n)$ cost of building the tree.
This is a significant improvement over exact partitioning and tree-building, which cost $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ respectively.

@@

We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that the tree is balanced.
This can realistically only happen when the data are uniformly distributed in the embedding space.
This represents the worst-case scenario for subsequent search algorithms because the LFD of the dataset will approach or equal the embedding dimension.
For real-world datasets with low LFD, we expect the tree to be highly unbalanced, reflecting the topological structure of the data.
When building trees for such datasets, their depths would be larger than $\log n$, which would seem to increase the cost of tree-building.
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of the tree would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

Given a dataset $\textbf{X}$, a distance function $f$ and the root cluster of the tree built using Algorithm~\ref{alg:methods:partition}, we can now formally pose the $k$-NN search problem.
Given a user specified integer $k$ and query $q$, $k$-NN search aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

@@

For Repeated $\rho$-NN search (Algorithm~\ref{alg:methods:repeated-rnn}), we begin by performing \textit{tree-search} with a search radius $r$ equal to the radius of the root cluster divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($root$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $root$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find
        \State $r \Leftarrow radius$ of the $root$ cluster
        \State $r \Leftarrow$ $\frac{r}{|root|}$
        \Loop
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
            \If{$Q \neq \emptyset$}
                \State \textbf{break}
            \EndIf
            \State $r \Leftarrow 2 \cdot r$
        \EndLoop
        \Loop
            \If{$\sum_{C \in Q} |C| >= k$}
                \State \textbf{break}
            \EndIf
            \State $\mu \Leftarrow |Q| \cdot \left( \sum_{C \in Q} LFD(C)^{-1} \right)^{-1}$
            \State $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)$
            \State $Q \Leftarrow$ tree-search($root$, $q$, $r$)
        \EndLoop
        \State $H \Leftarrow \text{leaf-search}(Q, q, r)$
        \State \textbf{sort} $H$ by non-decreasing distance to $q$
        \State \textbf{return} The first $k$ points in $H$
    \end{algorithmic}
\end{algorithm}

@@

\begin{minipage}{0.5125\textwidth}
    \begin{algorithm}[H]\small
        \caption{Breadth-First Sieve($root$, $q$, $k$)} % give the algorithm a caption
        \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
        \begin{algorithmic} % enter the algorithmic environment
            \Require $root$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find
            \State $c \Leftarrow$ \textit{center} of $root$
            \State $Q \Leftarrow$ \{ ($root$, $\delta^{+}_{root}$, $|root| - 1$), ($c$, $\delta_{root}$, 1) \}
            \Loop
                \If{$k = \sum_{(\_, \_, m) \in Q} m$}
                    \State \textbf{break}
                \EndIf
                \State $\tau \Leftarrow$ QuickSelect($Q$, $k$)
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \If{$\delta^{-}_{p} > \tau$}
                        \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \EndIf
                \EndFor
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \If{$p$ is a point}
                        \State \textbf{continue}
                    \ElsIf{$p$ is a leaf}
                        \For{$c \in p$}
                            \State $Q \Leftarrow Q \cup \{ (c, \delta_{c}, 1) \}$
                        \EndFor
                    \Else
                        \State $[L, R] \Leftarrow$ children of $p$
                        \State $l, r \Leftarrow$ centers of $L, R$
                        \State $Q \Leftarrow Q \cup \{ (L, \delta^{+}_{L}, |L| - 1), (l, \delta_{L}, 1) \}$
                        \State $Q \Leftarrow Q \cup \{ (R, \delta^{+}_{R}, |R| - 1), (l, \delta_{R}, 1) \}$
                    \EndIf
                \EndFor
            \EndLoop
            \State QuickSelect($Q$, $k$)
            \State \textbf{return} The first $k$ points in $Q$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.425\textwidth}
    \begin{algorithm}[H]\small
        \caption{Depth-First Sieve($root$, $q$, $k$)}
        \label{alg:methods:depth-first-sieve}
        \begin{algorithmic}
            \Require $root$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find

            \State{$Q \Leftarrow [root]$, a min-priority queue by $\delta^{-}$}
            \State{$H \Leftarrow []$, a max-priority queue by $\delta$}

@@

This algorithm (described in Algorithm~\ref{alg:methods:bredth-first-sieve}) performs a breadth-first traversal of the tree, pruning clusters by using a modified version of the QuickSelect algorithm~\cite{hoare1961algorithm} at each level.

@@

This algorithm (described in Algorithm~\ref{alg:methods:depth-first-sieve}) is quite similar to a depth-first traversal of the tree, using two priority queues to track clusters and hits, and to prioritize which branch of the tree to explore next.

Let $Q$ be a min-queue of clusters prioritized by $\delta^{-}$ and $H$ be a max-queue (with capacity $k$) of points prioritized by $\delta$.
$Q$ starts containing only the root cluster while $H$ starts empty.
So long as $H$ is not full or the top priority element in $H$ has $\delta$ greater than or equal to the top priority element in $Q$, we take the following steps:

@@

Note that this algorithm is not truly a depth-first traversal of the tree in the classical sense, because we use $Q$ to prioritize which branch of the tree we descend into.
Indeed, we expect this algorithm to often switch which branch of the tree is being explored at greater depth.

@@

We start with the assumption that the LFD near the query does not differ significantly from that of nearby clusters, and we consider leaf clusters with cardinalities near $k$.
Let $d$ be the LFD in this region.
Then the number of leaf-clusters in $Q$ is bounded above by $2d$, e.g.,\,we could have a cluster overlapping the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for tree-search, these leaf clusters would all come from different branches of the tree, and so tree-search looks at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity is $\mathcal{T} \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For leaf-search, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the asymptotic complexity is $\mathcal{L} \coloneqq \mathcal{O} \left( k \cdot \bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d \right)$.

The asymptotic complexity of Breadth-First Sieve is dominated by the QuickSelect algorithm to calculate $\tau$.
Since this method is log-linear in the length of $Q$, and $Q$ contains the clusters from tree-search and the points from leaf-search, we see that the asymptotic complexity is
\begin{equation}
    \mathcal{O} \Big( ( \mathcal{T} + \mathcal{L} ) \log ( \mathcal{T} + \mathcal{L} ) \Big)
    \label{eq:methods:breadth-first-sieve-complexity}
\end{equation}

For Depth-First Sieve, since we use two priority queues, the asymptotic complexity is dominated by the priority queue operations.
Thus, the complexity is
\begin{equation}
    \mathcal{O} \Big( \mathcal{T} \log \mathcal{T} + \mathcal{L} \log k \Big)
    \label{eq:methods:depth-first-sieve-complexity}
\end{equation}

@@

We perform some simple auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth, (e.g.,\,10), in the cluster tree as a query.
This gives us a small, representative sample of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{sec:methods:knn-search}.
We select the fastest algorithm over all the queries as the optimal algorithm for that dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

>>>>

We define a \textit{cluster} as a set of points with a \textit{center} and a \textit{radius}.
The center is the geometric median of the points in the cluster, i.e.,\,it is the point that minimizes the sum of distances to all other points in the cluster.
In cases where the cardinality of the cluster is large, we take a random subsample of $\sqrt{|C|}$ points and compute the geometric median of that subsample~\cite{ishaq2019clustered}.
The center, therefore, is one of the points in the cluster and is used as a representative of the cluster.
The radius is the maximum distance from the center to any point in the cluster.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
Note that clusters can have overlapping volumes and, in such cases, points in the overlapping volume are assigned to exactly one of the overlapping clusters.
As a consequence, a cluster can be a proper subset of the metric ball at the same center and radius, i.e.,\,$C(c, r) \subset B(c, r)$. We denote the cluster tree by $\mathcal{T}$ and the root by $\mathcal{R}$ (see Section~\ref{sec:methods:clustering:building-the-tree}).

@@

We start by performing a divisive hierarchical clustering on the dataset using CLAM to obtain a cluster tree $\mathcal{T}$.
The procedure is almost identical to that outlined in CHESS~\cite{ishaq2019clustered}, but with better selection of poles for partitioning (see Algorithm~\ref{alg:methods:partition}).
After building $\mathcal{T}$ with CLAM, we also perform a depth-first reordering of the dataset (see Section~\ref{sec:methods:clustering:depth-first-reordering}).

Given a cluster $C$ with $|C|$ points, we define its two children by the following process. We take a random subsample $S$ of $\sqrt{|C|}$ of $C$'s points, and compute pairwise distances between all points in $S$.
Using these distances, we compute the \textit{geometric median} of $S$; in other words, we find the point that minimizes the sum of distances to all other points in $S$.
We define the \textit{center} of $C$ to be this geometric median.

The \textit{radius} of $C$ is the maximum distance from the center to any other point in $C$.
The point that is responsible for that radius (i.e.,\,the furthest point from the center) is designated the \textit{left pole} and the point that is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster that are closer to the left pole than to the right pole, and the right child contains all points in the cluster that are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points that are equidistant from the two poles.
Starting from a root cluster $\mathcal{R}$ containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria (e.g., minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc).
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

Note that the tree $\mathcal{T}$ produced by Algorithm~\ref{alg:methods:partition} is not necessarily balanced.
In fact, for real datasets, we expect anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g.,\,in a $d$-dimensional hyper-cube.
In this sense, the imbalance in the tree is a feature, not a bug, as it reflects the underlying structure of the data.

@@

In CHESS~\cite{ishaq2019clustered}, each cluster stored a list of indices into the dataset.
This list was used to retrieve the clusters' points during search.
Although this approach allowed us to retrieve the points in constant time, its memory cost was prohibitively high.
With a dataset of cardinality $n$ and each cluster storing a list of indices for its points, we stored a total of $n$ indices at each depth in the tree $\mathcal{T}$.
Assuming $\mathcal{T}$ is balanced, and thus $\mathcal{O}(\log n)$ depth, this approach had a memory overhead of $\mathcal{O}(n \log n)$.
In this work, we introduce a new approach wherein, after building $\mathcal{T}$, we reorder the dataset so that points are stored in a depth-first order.
Then, within each cluster, we need only store its \textit{cardinality} and an \textit{offset} to access its points from the dataset, based on a useful invariant.
The root cluster $\mathcal{R}$ has an offset of zero and a cardinality equal to the number of points in the dataset.
A left child has the same offset its parent, and the corresponding right child has an offset equal to the left child's offset plus the left child's cardinality.
With no additional memory cost nor time cost for retrieving points during search, depth-first reordering offers the same time complexity as CHESS with $\mathcal{O}(n)$ space.

@@

The asymptotic complexity of Partition is the same as described in~\cite{ishaq2019clustered}.
By using an approximate partitioning with a $\sqrt{n}$ sample, we achieve $\mathcal{O}(n)$ cost of partitioning and $\mathcal{O}(n \log n)$ cost of building $\mathcal{T}$.
This is a significant improvement over exact partitioning and tree-building, which cost $\mathcal{O}(n^2)$ and $\mathcal{O}(n^2 \log n)$ respectively.

@@

We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that $\mathcal{T}$ is balanced.
This can realistically only happen when the data are uniformly distributed in the embedding space.
This represents the worst-case scenario for subsequent search algorithms because the LFD of the dataset will approach or equal the embedding dimension.
For real-world datasets with low LFD, we expect $\mathcal{T}$ to be highly unbalanced, reflecting the topological structure of the data.
When building trees for such datasets, their depths would be larger than $\log n$, which would seem to increase the cost of tree-building.
However, because of the large numbers of leaf clusters at shallow depths, each subsequent level of $\mathcal{T}$ would also have fewer points, and so the cost of building each subsequent level would be lower.
Any analysis of unbalanced trees would be highly dependent on the specific dataset, and so we do not provide one here.
In any case, the tree is built only once, so this cost is amortized over all search queries and other applications.

@@

Given a dataset $\textbf{X}$, a distance function $f$ and the root cluster $\mathcal{R}$ of the tree $\mathcal{T}$ (constructed by Algorithm~\ref{alg:methods:partition}), we can now formally pose the $k$-NN search problem.
Given a user specified integer $k$ and query $q$, $k$-NN search aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

@@

For Repeated $\rho$-NN search (Algorithm~\ref{alg:methods:repeated-rnn}), we begin by performing \textit{tree-search} with a search radius $r$ equal to the radius of the root $\mathcal{R}$ divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Repeated $\rho$-NN($\mathcal{R}$, $q$, $k$)} % give the algorithm a caption
    \label{alg:methods:repeated-rnn} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \Require $\mathcal{R}$, the root cluster
        \Require $q$, a query
        \Require $k$, the number of neighbors to find
        \State $r \Leftarrow radius$ of $\mathcal{R}$
        \State $r \Leftarrow$ $\frac{r}{|\mathcal{R}|}$
        \Loop
            \State $Q \Leftarrow$ tree-search($\mathcal{R}$, $q$, $r$)
            \If{$Q \neq \emptyset$}
                \State \textbf{break}
            \EndIf
            \State $r \Leftarrow 2 \cdot r$
        \EndLoop
        \Loop
            \If{$\sum_{C \in Q} |C| >= k$}
                \State \textbf{break}
            \EndIf
            \State $\mu \Leftarrow |Q| \cdot \left( \sum_{C \in Q} LFD(C)^{-1} \right)^{-1}$
            \State $r \Leftarrow r \cdot \min \left( 2, \left( {\frac{k}{\sum_{C \in Q} |C|}} \right)^{\mu^{-1}} \right)$
            \State $Q \Leftarrow$ tree-search($\mathcal{R}$, $q$, $r$)
        \EndLoop
        \State $H \Leftarrow \text{leaf-search}(Q, q, r)$
        \State \textbf{sort} $H$ by non-decreasing distance to $q$
        \State \textbf{return} The first $k$ points in $H$
    \end{algorithmic}
\end{algorithm}

@@

\begin{minipage}{0.5125\textwidth}
    \begin{algorithm}[H]\small
        \caption{Breadth-First Sieve($\mathcal{R}$, $q$, $k$)} % give the algorithm a caption
        \label{alg:methods:bredth-first-sieve} % and a label for \ref{} commands later in the document
        \begin{algorithmic} % enter the algorithmic environment
            \Require $\mathcal{R}$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find
            \State $c \Leftarrow$ \textit{center} of $\mathcal{R}$
            \State $Q \Leftarrow$ \{ ($\mathcal{R}$, $\delta^{+}_{\mathcal{R}}$, $|\mathcal{R}| - 1$), ($c$, $\delta_{\mathcal{R}}$, 1) \}
            \Loop
                \If{$k = \sum_{(\_, \_, m) \in Q} m$}
                    \State \textbf{break}
                \EndIf
                \State $\tau \Leftarrow$ QuickSelect($Q$, $k$)
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \If{$\delta^{-}_{p} > \tau$}
                        \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \EndIf
                \EndFor
                \For{$(p, \delta^{+}_{p}, m) \in Q$}
                    \State $Q \Leftarrow Q \setminus \{ (p, \delta^{+}_{p}, m) \}$
                    \If{$p$ is a point}
                        \State \textbf{continue}
                    \ElsIf{$p$ is a leaf}
                        \For{$c \in p$}
                            \State $Q \Leftarrow Q \cup \{ (c, \delta_{c}, 1) \}$
                        \EndFor
                    \Else
                        \State $[L, R] \Leftarrow$ children of $p$
                        \State $l, r \Leftarrow$ centers of $L, R$
                        \State $Q \Leftarrow Q \cup \{ (L, \delta^{+}_{L}, |L| - 1), (l, \delta_{L}, 1) \}$
                        \State $Q \Leftarrow Q \cup \{ (R, \delta^{+}_{R}, |R| - 1), (l, \delta_{R}, 1) \}$
                    \EndIf
                \EndFor
            \EndLoop
            \State QuickSelect($Q$, $k$)
            \State \textbf{return} The first $k$ points in $Q$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{0.425\textwidth}
    \begin{algorithm}[H]\small
        \caption{Depth-First Sieve($\mathcal{R}$, $q$, $k$)}
        \label{alg:methods:depth-first-sieve}
        \begin{algorithmic}
            \Require $\mathcal{R}$, the root cluster
            \Require $q$, a query
            \Require $k$, the number of neighbors to find

            \State{$Q \Leftarrow [\mathcal{R}]$, a min-priority queue by $\delta^{-}$}
            \State{$H \Leftarrow []$, a max-priority queue by $\delta$}

@@

This algorithm (described in Algorithm~\ref{alg:methods:bredth-first-sieve}) performs a breadth-first traversal of the tree $\mathcal{T}$, pruning clusters by using a modified version of the QuickSelect algorithm~\cite{hoare1961algorithm} at each level.

@@

This algorithm (described in Algorithm~\ref{alg:methods:depth-first-sieve}) is quite similar to a depth-first traversal of the tree $\mathcal{T}$, using two priority queues to track clusters and hits, and to prioritize which branch of $\mathcal{T}$ to explore next.

Let $Q$ be a min-queue of clusters prioritized by $\delta^{-}$ and $H$ be a max-queue (with capacity $k$) of points prioritized by $\delta$.
$Q$ starts containing only the root cluster $\mathcal{R}$ while $H$ starts empty.
So long as $H$ is not full or the top priority element in $H$ has $\delta$ greater than or equal to the top priority element in $Q$, we take the following steps:

@@

Note that this algorithm is not truly a depth-first traversal of $\mathcal{T}$ in the classical sense, because we use $Q$ to prioritize which branch of $\mathcal{T}$ we descend into.
Indeed, we expect this algorithm to often switch which branch of $\mathcal{T}$ is being explored at greater depth.

@@

We start with the assumption that the LFD near the query does not differ significantly from that of nearby clusters, and we consider leaf clusters with cardinalities near $k$.
Let $d$ be the LFD in this region.
Then the number of leaf-clusters in $Q$ is bounded above by $2d$, e.g.,\,we could have a cluster overlapping the query ball at each end of each of $\lceil d \rceil$ mutually-orthogonal axes.
In the worst-case scenario for tree-search, these leaf clusters would all come from different branches of the tree, and so tree-search looks at $2 \cdot \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X)$ clusters.
Thus, the asymptotic complexity is $T \coloneqq \mathcal{O} \big( \lceil d \rceil \cdot \log \mathcal{N}_{\hat{r}}(X) \big)$.
For leaf-search, the output size and scaling factor are the same as in Repeated $\rho$-NN, and so the asymptotic complexity is $L \coloneqq \mathcal{O} \left( k \cdot \bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d \right)$.

The asymptotic complexity of Breadth-First Sieve is dominated by the QuickSelect algorithm to calculate $\tau$.
Since this method is log-linear in the length of $Q$, and $Q$ contains the clusters from tree-search and the points from leaf-search, we see that the asymptotic complexity is
\begin{equation}
    \mathcal{O} \Big( (T + L ) \log (T + L ) \Big).
    \label{eq:methods:breadth-first-sieve-complexity}
\end{equation}

For Depth-First Sieve, since we use two priority queues, the asymptotic complexity is dominated by the priority queue operations.
Thus, the complexity is
\begin{equation}
    \mathcal{O} \Big( T \log T + L \log k \Big).
    \label{eq:methods:depth-first-sieve-complexity}
\end{equation}

@@

We perform some simple auto-tuning to select the optimal $k$-NN algorithm to use with a given dataset.
We start by taking the center of every cluster at a low depth (e.g.,\,10) in $\mathcal{T}$ as a query.
This gives us a small, representative sample of the dataset.
Using these clusters' centers as queries, and a user-specified value of $k$, we record the time taken for $k$-NN search on the sample using each of the three algorithms described in Section~\ref{sec:methods:knn-search}.
We select the fastest algorithm over all the queries as the optimal algorithm for that dataset and value of $k$.
Note that even though we select the optimal algorithm based on use with some user-specified value of $k$, we still allow search with any value of $k$.

### 2025-01-10 sections/4-results.tex
added notation for tree and root

In contrast with Figure~\ref{fig:results:sift-lfd}, Figure~\ref{fig:results:random-lfd} shows the LFD by depth for a random dataset with the same cardinality and dimensionality as those of Sift.
We observe that the LFD starts as high as 20 (for all data) at depth 0.
As shown by the needlepoint shape of the plot, the LFD for all clusters has a very narrow spread from depth 0 through depth 5.
After depth 5, we begin to see a wider spread between the maximum and minimum LFD at each depth.
All percentile lines seem to decrease linearly with depth.
The LFD of approximately 20 at depth 0 (i.e.,\,the root cluster) is what we expect for this random dataset.
To elaborate, the distribution of points in such a dataset should reflect the curse of dimensionality, i.e.,\,the fact that in high dimensional spaces, the minimum and maximum pairwise distances between any two points are approximately equal.
As a result, the root cluster's radius $r$, which reflects the maximum distance between the center $c$ and any other point, should not differ significantly from the distance between the center and its closest point.
A consequence of this is that, with high probability, for every point in the root cluster, its distance from $c$ is greater than $\tfrac{r}{2}$;
in other words, $B_X(c, \tfrac{r}{2})$ contains only $c$ while $B_X(c, r)$ contains the entire dataset.
Given our definition of LFD in Equation~\ref{eq:methods:lfd-original}, this means that the LFD of the root cluster is approximately $\log_2(\frac{|X|}{1}) = \log_2(1,000,000) \approx 20$, which is what we observe in Figure~\ref{fig:results:random-lfd}.

>>>>

In contrast with Figure~\ref{fig:results:sift-lfd}, Figure~\ref{fig:results:random-lfd} shows the LFD by depth for a random dataset with the same cardinality and dimensionality as those of Sift.
We observe that the LFD starts as high as 20 (for all data) at depth 0.
As shown by the needlepoint shape of the plot, the LFD for all clusters has a very narrow spread from depth 0 through depth 5.
After depth 5, we begin to see a wider spread between the maximum and minimum LFD at each depth.
All percentile lines seem to decrease linearly with depth.
The LFD of approximately 20 at depth 0 (i.e.,\,the root cluster $\mathcal{R}$) is what we expect for this random dataset.
To elaborate, the distribution of points in such a dataset should reflect the curse of dimensionality, i.e.,\,the fact that in high dimensional spaces, the minimum and maximum pairwise distances between any two points are approximately equal.
As a result, $\mathcal{R}$'s radius $r$, which reflects the maximum distance between the center $c$ and any other point, should not differ significantly from the distance between the center and its closest point.
A consequence of this is that, with high probability, for every point in $\mathcal{R}$, its distance from $c$ is greater than $\tfrac{r}{2}$;
in other words, $B_X(c, \tfrac{r}{2})$ contains only $c$ while $B_X(c, r)$ contains the entire dataset.
Given our definition of LFD in Equation~\ref{eq:methods:lfd-original}, this means that the LFD of $\mathcal{R}$ is approximately $\log_2(\frac{|X|}{1}) = \log_2(1,000,000) \approx 20$, which is what we observe in Figure~\ref{fig:results:random-lfd}.

### 2025-01-10 sections/5-discussion.tex
added notation for tree and root

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{\mathcal{T}}{\lceil d \rceil} + \mathcal{L})$, where $\mathcal{T}$ is the time complexity of tree-search and $\mathcal{L}$ is the time complexity of leaf-search.
Even though Repeated $\rho$-NN has the lowest time complexity (compare to $\mathcal{O}(\mathcal{T}\log{(\mathcal{T}+\mathcal{L})} + \mathcal{L}\log{(\mathcal{T}+\mathcal{L})})$ for Breadth-First Sieve and $\mathcal{O}(\mathcal{T}\log{\mathcal{T}} + \mathcal{L}\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically.
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in Equation~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation.
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD,
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase in (Equation)~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.
This suggests that rather than using the reciprocal of the harmonic mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may achieve better results with a mean that is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly ``overshoots'' $\rho_k$ and to improve upon the factor of radial increase in~\ref{eq:methods:repeated-rnn-factor} so that this occurs less frequently and with less severity.

>>>>

When written with the same notation as used in Section~\ref{sec:methods:knn-search:complexity-of-sieve-methods}, we see that the time complexity of Repeated $\rho$-NN is $\mathcal{O}(\tfrac{T}{\lceil d \rceil} + L)$, where $T$ is the time complexity of tree-search and $L$ is the time complexity of leaf-search.
Even though Repeated $\rho$-NN has the least time cost (compare to $\mathcal{O}(T\log{(T+L)} + L\log{(T+L)})$ for Breadth-First Sieve and $\mathcal{O}(T\log{T} + L\log{k})$ for Depth-First Sieve) of the three CAKES algorithms, it is not always the fastest algorithm empirically.
We believe that some of this discrepancy can be explained by the fact that if Repeated $\rho$-NN significantly ``overshoots'' the correct radius for $k$ hits ($\rho_k$) during tree-search, leaf-search will require looking at more points, rendering the true scaling factor higher than that in Equation~\ref{eq:methods:repeated-rnn-complexity}.
This overshot can occur when the LFDs of clusters near the query are not concentrated around their expectation.
For example, if most clusters near the query have very low LFD except for one anomaly with very high LFD,
the harmonic mean LFD $\mu$ can still be low, so the factor of radial increase in (Equation)~\ref{eq:methods:repeated-rnn-factor} may be much larger than necessary for guaranteeing $k$ hits.
This suggests that rather than using the reciprocal of the harmonic mean LFD in~\ref{eq:methods:repeated-rnn-factor}, we may achieve better results with a mean that is more sensitive to high outliers, such as the geometric mean.
We leave it as an avenue for future work to characterize when Repeated $\rho$-NN significantly ``overshoots'' $\rho_k$ and to improve upon the factor of radial increase in~\ref{eq:methods:repeated-rnn-factor} so that this occurs less frequently and with less severity.
